A hazards approach to the biometric analysis of infant mortality

### 2023-03-15 Slides/slides.Rmd
more info

>>>>

## Väterbotten 1861-1890

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/bioume6190-1.pdf}

## Västerbotten 1921--1950

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/bioume9120-1.pdf}

## Comparison

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/both-1.pdf}

## Cumulative hazards, neonatal mortality, Västerbotten 1861-1950

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/enof-1.pdf}

## Västerbotten, Bourgeois-Pichat's method

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/first-1.pdf}

## Västerbotten, hazards method

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/second-1.pdf}

## Västerbotten 1921--1950

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/lowumee-1.pdf}

## Exponential fits to post-neonatal data, Västerbotten

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/postneo-1.pdf}

## Scania 1711--1800

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/scahigh-1.pdf}

##

\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/skum-1.pdf}

### 2023-03-15 paper2.Rmd
more info

---
title: "A hazards approach to the biometric analysis of infant mortality"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: bb.bib
titlepage: no
biblio-style: apalike
setcitestyle: authoryear
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{natbib}
- \setcitestyle{authoryear}
- \usepackage{a4wide}
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
##- \subsectionfont{\normalfont\normalfont\itshape}
abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is suggested. In the original model, cumulative mortality in the last eleven months of infancy is assumed to follow a uniform distribution given a log-cube transformation of age. Instead, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying. Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality."
---

@@

Lois Bourgeois-Pichat saw the first year of life not only as the period when mortality is highest, but also the period when improvements should be possible 
[@bp51a; @bp51b; @bp52]. He argued that infant deaths should be divided into two categories, endogenous and exogenous deaths. Endogenous deaths are, by definition, due to inherited factors, or acquired during gestation or delivery. They typically occur at the initial period of life, though sometimes much later and include congenital debility, prematurity, malformations, and disease in early life. Exogenous infant mortality, which Bourgeois-Pichat regards as accidental, are deaths for which the society must hold itself responsible and to reduce endogenous mortality, medical intervention is essential [@bp52]. Since it is, based on causes of deaths, difficult to distinguish between endogenous and exogenous factors, either because they are inaccurate or difficult to make due to the diseases themself, or the information do not exist, Bourgeois-Pichat offered a solution to the problem.

@@

Applying this method, Bourgeois-Pichat also finds deviations from the linear death pattern after the first month. To take one example, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12]. To take another example, in the case of Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8]. A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec. @knoki77 argues that breastfeeding patterns have a strong impact on the linearity. The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].

Bourgeois-Pichat's biometric method has also been used to identify problems with data recording [@wea77; @weasr81]. While a high level of endogenous mortality cannot be used as a criterion for high data quality, the opposite holds true. A very low level of endogenous mortality indeed indicates data problems, just like large deviations from the normal sex ratio at birth do. These two criteria are, in fact, often used in historical studies to evaluate data recording (see Bengtsson et al 2004). They become even more precise if they are applied to different social strata, since under-recording of early deaths often have a social gradient, possibly due to the costs involved in a burial [@tb99; @tbcl99].

We suggest a variation of the Bourgeois-Pichat biometric analysis of infant mortality. Instead of assuming that the cumulative mortality in the last eleven months of infancy follows a uniform distribution, given a log-cube transformation of age, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. This assumption is more satisfactory from a theoretical point of view, since it models the conditional probability of dying. The growth in birth weight also follows this distribution [@bp51a; @bp51b]. In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies. The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.1 In addition, we give examples from eighteenth Sweden, where the curve do not follow the uniform distribution during the last eleven month of infancy, despite high levels of mortality in the first month of life.

@@

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
we use a data set from northern Sweden, covering the years 1861--1950.
```{r readdata}
infant <- readRDS("Rdata/infant.rds")
infant$period <- cut(infant$birthdate, 
                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
                                      "1920-12-31", "1951-01-01")),
                   labels = c("1861-1890", "1891-1920", "1921-1950"))
infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "icd.chapter")]
ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```
For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).

```{r first, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, Bourgeois-Pichat method."}
source("R/drawBP.R")
oldpar <- par(mfrow = c(1,2))
##
r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
g1 <- infant[infant$period == "1861-1890", ]
r1 <- eha::risksets(r1, members = FALSE)
drawBP(r1, main = "1861-1890", ylim = c(0, 0.1))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
g3 <- infant[infant$period == "1921-1950", ]
r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
r3 <- eha::risksets(r3, members = FALSE)
drawBP(r3, main = "1921-1950", ylim = c(0, 0.1))
par(oldpar)
```

>>>>

---
title: "A hazards approach to the biometric analysis of infant mortality"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: bb.bib
titlepage: no
biblio-style: apalike
setcitestyle: authoryear
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{natbib}
- \setcitestyle{authoryear}
- \usepackage{a4wide}
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
##- \subsectionfont{\normalfont\normalfont\itshape}
abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is suggested. In the original model, cumulative mortality in the last eleven months of infancy is assumed to follow a uniform distribution given a log-cube transformation of age. Instead, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying. Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality. We use data from northern and southern Sweden for the period 1710 to 1950 to compare the two methods."
---

@@

Lois Bourgeois-Pichat saw the first year of life not only as the period when mortality is highest, but also the period when improvements should be possible 
[@bp51a; @bp51b; @bp52]. He argued that infant deaths should be divided into two categories, endogenous and exogenous deaths. Endogenous deaths are, by definition, due to inherited factors, or acquired during gestation or delivery. They typically occur at the initial period of life, though sometimes much later and include congenital debility, prematurity, malformations, and disease in early life. Exogenous infant mortality, which Bourgeois-Pichat regards as accidental, are deaths for which the society must hold itself responsible and to reduce endogenous mortality, medical intervention is essential [@bp52]. Since it is, based on causes of deaths, difficult to distinguish between endogenous and exogenous factors, either because they are inaccurate or difficult to make due to the diseases themselves, or the information do not exist, Bourgeois-Pichat offered a solution to the problem.

@@

Bourgeois-Pichat also found deviations from the linear death pattern after the first month, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12], and in the case of Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8]. A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec. @knoki77 argues that breastfeeding patterns have a strong impact on the linearity, resulting in increasing mortality after weaning. The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].

Bourgeois-Pichat's biometric method has also been used to identify problems with data recording [@wea77; @weasr81]. While a high level of endogenous mortality cannot be used as a criterion for high data quality, the opposite holds true. A very low level of endogenous mortality indeed indicates data problems, just like large deviations from the normal sex ratio at birth do. These two criteria are, in fact, often used in historical studies to evaluate data quality [@bcl04]. They become even more precise if they are applied to different social strata, since under-recording of early deaths often have a social gradient, possibly due to the costs involved in a burial [@tb99; @tbcl99].

We suggest a variation of the Bourgeois-Pichat biometric analysis of infant mortality. Instead of assuming that the cumulative mortality in the last eleven months of infancy follows a uniform distribution, given a log-cube transformation of age, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. This assumption is more satisfactory from a theoretical point of view, since it models the conditional probability of dying. The growth in birth weight also follows this distribution [@bp51a; @bp51b]. In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies. The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.1 In addition, we give examples from eighteenth Sweden, where the curve is concave during the last eleven month of infancy, despite high levels of mortality in the first month of life.

@@

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
we use a data set from northern Sweden, covering the years 1861--1950.
```{r readdata}
infant <- readRDS("Rdata/infant.rds")
infant$month <- months(infant$birthdate)
infant$period <- cut(infant$birthdate, 
                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
                                      "1920-12-31", "1951-01-01")),
                   labels = c("1861-1890", "1891-1920", "1921-1950"))
infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "month")]
ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```
For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).

```{r first, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, Bourgeois-Pichat method."}
source("R/drawBP.R")
##infant <- infant[infant$month %in% c("juni", "juli", "augusti"), ]
##ginfant <- ginfant[ginfant$month %in% c("juni", "juli",  "augusti"), ]
oldpar <- par(mfrow = c(1,2))
##
r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
g1 <- infant[infant$period == "1861-1890", ]
r1 <- eha::risksets(r1, members = FALSE)
drawBP(r1, main = "1861-1890", ylim = c(0, 0.1))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
g3 <- infant[infant$period == "1921-1950", ]
r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
r3 <- eha::risksets(r3, members = FALSE)
drawBP(r3, main = "1921-1950", ylim = c(0, 0.1))
par(oldpar)
```
